---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesGP

<!-- badges: start -->
<!-- badges: end -->

The goal of the `BayesGP` package is to efficiently implement model-based smoothing with flexible GP priors, within a variety of Bayesian hierarchical models.

## Installation

You can install the development version of BayesGP from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("https://github.com/Bayes-GP/BayesGP/tree/development")
```

## Example: sGP

In this example, we will use the `lynx` dataset as an example, which can be accessed directly from R. Let's load the dataset and visualize it:

```{r data}
library(BayesGP)
data <- data.frame(year = seq(1821, 1934, by = 1), y = as.numeric(lynx))
data$x <- data$year - min(data$year)
plot(lynx)
```

Based on a visual examination of the dataset, we can observe an obvious 10-year quasi-periodic behavior in the lynx count with evolving amplitudes over time. Therefore, we will consider fitting the following model:

$$
\begin{equation}
    \begin{aligned}
        y_i|\lambda_i &\sim \text{Poisson}(\lambda_i) ,\\
        \log(\lambda_i) &= \eta_i = \beta_0 + g(x_i) + \xi_i,\\
        g &\sim \text{sGP} \bigg(\alpha = \frac{2\pi}{10}, \sigma\bigg),\\
        \xi_i &\sim N(0,\sigma_\xi).
    \end{aligned}
\end{equation}
$$
Here, each $y_i$ represents the lynx count, $x_i$ represents the number of years since 1821, and $\xi_i$ is an observation-level random intercept to account for overdispersion effect.


To specify the priors for the sGP boundary conditions and the intercept parameter, we assume independent normal priors with mean 0 and variance 1000.
For the overdispersion parameter $\sigma_\xi$, we assign an exponential prior with $P(\sigma_\xi > 1) = 0.01$.


To determine the prior for the standard deviation parameter $\sigma$ of the sGP, we employ the concept of predictive standard deviation (PSD). We start with an exponential prior on the 50 years PSD:
$$P(\sigma(50)>1) = 0.01.$$


To make inference of the quasi-periodic seasonal effect through the seasonal-B spline approach, we make use of the `model_fit` function from `BayesGP`. 
The prior information of the sGP component should be specified in `f(.)`.
The argument `k = 30` will setup 30 equally spaced seasonal B-spline basis functions for the approximation.
A higher value of `k` will increase the approximation accuracy of the seasonal B-spline approximation, at the cost of longer runtime.

By default, the posterior is obtained through an approximation using the adaptive quadrature method described in the main paper. 
However, exact method through MCMC sampling is also possible, by specifying `method = "MCMC"` in `model_fit`.

```{r}
mod <- model_fit(formula = y ~ f(x, model = "sgp", period = 10, k = 30,
                                 sd.prior = list(param = list(u = 1, alpha = 0.01), h = 50),
                                 boundary.prior = list(prec = 0.001)) +
                   f(year, model = "iid", sd.prior = list(param = list(u = 1, alpha = 0.01))),
                 family = "Poisson", data = data, method = "aghq")
```


The posterior summary of the fitted model can be examined through `summary` or `plot`:

```{r}
summary(mod)
```


```{r}
plot(mod)
```




## Example: IWP

This is a basic example which shows you how to use `BayesGP` to fit and
analyze some models, we consider the following data set of COVID-19 mortality in Canada, which is available in the package:

```{r example}
library(BayesGP)
## basic example code
head(covid_canada)
```

We can fit a model with $\text{IWP}_3(\sigma)$ prior using the function `model_fit`:
```{r warning=FALSE}
fit_result <- model_fit(new_deaths ~ weekdays1 + weekdays2 + weekdays3 + weekdays4 + weekdays5 + weekdays6 +
                          f(smoothing_var = t, model = "IWP", order = 3, k = 30), 
                        data = covid_canada, method = "aghq", family = "Poisson")
```

We can take a look at the posterior summary of this model:
```{r}
summary(fit_result)
```

We can also see the inferred function $f$:
```{r warning=FALSE}
plot(fit_result)
```


We can use the `predict` function to obtain the posterior summary of $f$ or its derivative at `new_data`. 

For the function $f$:
```{r warning=FALSE}
library(tidyverse)
predict_f <- predict(fit_result, variable = "t", newdata = data.frame(t = seq(from = 605, to = 617, by = 0.1)))
predict_f %>% ggplot(aes(x = t)) + geom_line(aes(y = mean), lty = "solid") +
  geom_line(aes(y = `q0.025`), lty = "dashed") +
  geom_line(aes(y = `q0.975`), lty = "dashed") +
  theme_classic()
```


For the first derivative:
```{r warning=FALSE}
predict_f1st <- predict(fit_result, variable = "t", newdata = data.frame(t = seq(from = 605, to = 617, by = 0.1)), deriv = 1)
predict_f1st %>% ggplot(aes(x = t)) + geom_line(aes(y = mean), lty = "solid") +
  geom_line(aes(y = `q0.025`), lty = "dashed") +
  geom_line(aes(y = `q0.975`), lty = "dashed") +
  theme_classic()
```

For the second derivative:
```{r warning=FALSE}
predict_f2nd <- predict(fit_result, variable = "t", newdata = data.frame(t = seq(from = 605, to = 617, by = 0.1)), deriv = 2)
predict_f2nd %>% ggplot(aes(x = t)) + geom_line(aes(y = mean), lty = "solid") +
  geom_line(aes(y = `q0.025`), lty = "dashed") +
  geom_line(aes(y = `q0.975`), lty = "dashed") +
  theme_classic()
```



